Reproducible Data Assignment 2
==============================

The goal of this assignment is to explore the NOAA Storm Database and answer two questions about severe weather events:

1. Across the United States, which types of events (as indicated in the *evtype* variable) are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

# Data Processing


## Load the data

Download the dataset and load it into a dataframe.

```{r}
library(plyr)
library(ggplot2)
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
destfile <- "stormdata.csv.bz2"
if (!file.exists(destfile)) {
        download.file(url = url, destfile = destfile, method = "curl")
}

data <- read.csv(destfile)

```

## Preprocess the data

Tidy the column names by making them lowercase. 

The "evtype" column has numerous issues:
1. Misspellings
2. Mix of lowercase and uppercase values
3. Mix of plural and singular names
4. Mix of specific events such as "hurricane <name>" mixed with event types

The following code corrects the issues with "evtype" values.

```{r}
colnames(data) <- tolower(colnames(data))

## Subset of columns for mapping events to health
healthColumns <- c("evtype","fatalities", "injuries")
healthData <- data[healthColumns]

## evtype names are both UPPER and lowercase make them all lowercase for grouping
healthData$evtype <- trimws(tolower(healthData$evtype), c("both"))

## Some events are plural, convert plural to singular events
healthData$evtype <- ifelse (grepl(".*s$", healthData$evtype),
                             substr(healthData$evtype, 1, nchar(healthData$evtype) - 1), 
                             healthData$evtype)

## Convert all versions of flood to flood
healthData$evtype <- ifelse (grepl("flood", healthData$evtype), 
                             "flood",
                             healthData$evtype)

## Convert all versions of hurriance to huricane
healthData$evtype <- ifelse (grepl("hurricane", healthData$evtype), 
                             "hurricane", 
                             healthData$evtype)

## Convert all versions of thunderstorm to thunderstorm
healthData$evtype <- ifelse (grepl("thunder", healthData$evtype), 
                             "thunderstorm", 
                             healthData$evtype)

## Convert all versions of ic* to ice
healthData$evtype <- ifelse (grepl("^ic", healthData$evtype), 
                             "ice", 
                             healthData$evtype)

## Convert all versions of wint* to winter weather
healthData$evtype <- ifelse (grepl("^wint", healthData$evtype), 
                             "winter weather", 
                             healthData$evtype)

## Convert all versions of wint* to winter weather
healthData$evtype <- ifelse (grepl("snow", healthData$evtype), 
                             "snow", 
                             healthData$evtype)
```
# Results

Which types of events (as indicated in the *evtype* variable) are most harmful with respect to population health?

We have data which gives us counts for injuries and fatalities for each event type. We'll group by evtype across all observations and sum injuries and fatalities.

Next we'll create a new value which is the sum of injuries + fatalities for each evtype.
```{r}
healthSummary <- ddply(healthData, .(evtype), numcolwise(sum))
healthSummary$injury_death <- healthSummary$fatalities + healthSummary$injuries
healthSummary <- subset(healthSummary, healthSummary$injury_death > 0)
healthSummary <- healthSummary[order(healthSummary$injury_death,decreasing = T),]


```
### 97% of all injuries are caused by 20 event types. Let's focus on those top 20 events.
```{r}
## Get the top 20
top20 <- healthSummary[1:20,]
sum(top20$injury_death) / sum(healthSummary$injury_death) * 100
```
### Most fatalities and injuries are from Tornados

```{r}
top20$events <- factor(top20$evtype, levels=top20$evtype, ordered=TRUE)
g <- ggplot(top20, aes(x=as.factor(events),y=injury_death)) +
        labs(x = "Event Type") + 
        labs(y = "Fatalities and Injuries") +
        labs(title = "Fatalities and Injuries by Weather Event") +
        geom_bar(stat = "Identity") + 
        theme(axis.text.x = element_text(angle = 90, hjust = 1))
print(g)
```

